Evolution of the AGN UV luminosity function from redshift 7.5

### 2018-03-30 Paper/paper.tex
Update text

\author[Kulkarni et al.]{{Girish Kulkarni$^{1}$\thanks{Email:
      kulkarni@ast.cam.ac.uk}, G\'abor Worseck$^{2}$ and Joseph
    F.~Hennawi$^{3}$} \\ $^1$Institute of Astronomy and Kavli
  Institute of Cosmology, University of Cambridge, Madingley Road,
  Cambridge CB3 0HA, UK \\ $^2$Max Planck Institute for Astronomy,
  K\"onigstuhl 17, D-69117 Heidelberg, Germany\\ $^3$Department of
  Physics, Broida Hall, UC Santa Barbara, Santa Barbara, CA 93106-9530
  USA}
% Update Gabor's address to Potsdam
\date{Accepted ---. Received ---; in original form ---}

@@

\begin{figure}
  \begin{center}
    \includegraphics[width=\columnwidth,keepaspectratio]{q.pdf}
  \end{center}
  \caption{Evolution of the \HeIII\ ionization fraction in our model,
    when the qso luminosity function is integrated down to
    $M_{1450}=-21$ (black curve) and when it is extrapolated and
    integrated down to $M_{1450}=-18$ (brown curve).  Other curves
    show models of \citet[solid grey]{2012ApJ...746..125H},
    \citet[dashed grey]{2015ApJ...813L...8M}, \citet[light blue and
      shaded region]{2016ApJ...828...90L},
    \citet[blue]{2018arXiv180104931P}. \gk{Do we need to add any more
      models from literature?}}
\end{figure}

We now consider the implications for hydrogen and helium reionization
for the ionizing photon emissivity derived above.  This requires that
we extrapolate the emissivity of Equation XX to higher redshifts.
Unfortunately, the volume emissivity in our model cannot be
extrapolated to higher redshifts.  The model of Hopkins suffered with
the same problem.  \gk{Do we need to show two cases here?  One in
  which we extrapolate the emissivity and another in which we stop
  where data end.  There is perhaps a third possibility in which use
  the current emissivity model upto where data end.}  The evolution of
the volume averaged ionized fraction of hydrogen is given by \gk{Is
  this really volume-weighted?}  \citep{2012ApJ...746..125H}
\begin{equation}
{dQ_\nHII\over dt}={\dot n_{\rm ion}\over \langle n_\nH \rangle} -{Q_\nHII\over
\langle t_{\rm rec}\rangle},
\label{eq:qdot}
\end{equation}
and the corresponding equation doubly ionized Helium is
\begin{equation}
{dQ_\nHeIII\over dt}={\dot n_{\rm ion,4}\over \langle n_\nHe \rangle} -{Q_\nHeIII\over
\langle t_{\rm rec,He}\rangle}.
\label{eq:qHedot}
\end{equation}
We assume that the Helium is singly reionized together with hydrogen.
In order to get the 4~Ry emissivity of quasars we use the SED from
equation XX.  The recombination times are given by
\begin{equation}
\langle t_{\rm rec}\rangle=[\chi \langle n\rangle \alpha_B\,C]^{-1},
\label{eq:trec}
\end{equation}
where $n$ is the volume averaged density of hydrogen and helium,
respectively.  We assume that hydrogen and helium have similar
clumping factors, given by \citep{2015ApJ...813L...8M}
\begin{equation}
  C = 2.8\left(\frac{1+z}{6}\right)^{-1.1}.
\end{equation}
For the helium recombination coeffienct, we use the fitting function
provided by \citet{1997MNRAS.292...27H}.  For hydrogen recombination
coefficient, we use the fitting function provided by
\citet{2011piim.book.....D}.

The resultant reionization histories of hydrogen and helium are shown
in Figure XX.  We also show the reionization histories from
\citep{2012ApJ...746..125H} and \citep{2015ApJ...813L...8M} for
comparison.  As expected from the emissivity result of the previous
section, quasars cannot single handedly reionize hydrogen.  However
they can reionise \HeII.  The Helium reionization completes at
redshift $z=3.5$--$3.7$ depending on the luminosity down to which you
integrate the quasar luminosity function.  This is consistent with the
Helium \lya opacity measurements of \citet{2016ApJ...825..144W}.
Discuss Mitra et al. 2018 and Khaire and Srianand 2018.  See Gabor's
email of 14 March 13:18.

@@

We thank Eilat Glikman, Linhua Jiang, Nobunari Kashi\-kawa, Ian
McGreer, Nick Ross, Chris Willott, and Jinyi Yang for sharing data and
especially for sharing their quasar selection functions.  It is a
pleasure to acknowledge useful discussions with James Aird, Eduardo
Ba\~nados, Manda Banerji, Tirthankar Roy Choudhury, George Efstathiou,
Xiaohui Fan, Andrea Ferrara, Prakash Gaikwad, Martin Haehnelt, Paul
Hewett, David Hogg, Vikram Khaire, Sergey Koposov, Donald Lynden-Bell,
Roberto Maiolino, Richard McMahon, Daniel Mortlock, Ewald Puchwein,
Gordon Richards, Alberto Rorai, Bram Venemans and Stephen Warren.  GK
acknowledges support from ERC Advanced Grant 320596 `The Emergence of
Structure During the Epoch of Reionization'. \gk{Joe, do you need to
  add anything to acknowledgements?}

@@

\section{Comparison with Giallongo sample}
\label{sec:conv}

In our analysis above, we did not include the 22 low-luminosity
($M_{1450}>-22.6$) AGN between redshifts $z=4.1$ and $6.3$ reported by
\citet{2015AA...578A..83G} using a novel X-ray/NIR selection
criterion.  This was because we wanted to restrict our sample to
quasars that have robust spectroscopic redshift determinations.  In
their work, \citet{2015AA...578A..83G} found a steep faint-end of the
luminosity function at these redshifts.  With this luminosity function
they obtained a large value of hydrogen ionizing emissivity that could
explain the \lya data singlehandedly.  In this section, we compare our
results with \citet{2015AA...578A..83G}.  The black points in Figure
XX show the parameters of the double power law luminosity function
when the \citet{2015AA...578A..83G} sample is added to our analysis.
The parameter values do not change significantly.  The
\citet{2015AA...578A..83G} sample is thus consistent with the double
power law luminosity function preferred by other data at comparable
redshifts.  However, as we see in Figure XX, the total emissivity in
our model is still significantly lower than that obtained by
\citet{2015AA...578A..83G}.  Figure XX provides an explanation for
this. The double power law fits favoured by \citet{2015AA...578A..83G}
are quite different from our fits.  The character luminosity $M_*$
favoured by \citet{2015AA...578A..83G} is much fainter ($\sim -23$ at
$z = 5$) than that resulting out of our analysis ($\sim -29$ at $z =
5$).  Figure XX suggests that this is because of the SDSS Stripe 82
data \gk{citation} and because \citet{2015AA...578A..83G} ignore the
Kashikawa and Willott samples \gk{citation}.  It is possible that our
corrections to the Glikman data \gk{citation} also plays a role in
this.  In sum, the differnce in our emissivities is not due to lack of
faint quasars but because of the difference with Giallongo at
intermediate luminosities.  However, our double power law fits are
virtually identical to those shown in Figure~\ref{fig:mosaic}.

\begin{figure*}
  \begin{center}
    \includegraphics[width=0.7\textwidth]{evolution_g.pdf}
  \end{center}
  \caption{Effect of the 19 qsos reported by
    \citet{2015AA...578A..83G} on the best-fit double power law
    luminosity function parameters in redshift bins from $z=0$ to $7$.
    Black points show parameter values from Figure~\ref{fig:evoln}.
    Red open circles show the parameter values obtained when the
    sample of \citet{2015AA...578A..83G} is added to the analysis.  In
    both cases, vertical error bars show one-sigma (68.26\%)
    uncertainties, whereas horizontal error bars show widths of the
    redshift bins.}
\end{figure*}

\begin{figure*}
  \begin{center}
    \includegraphics[width=\textwidth]{giallongo_compare.pdf}
  \end{center}
  \caption{Luminosity functions in three redshift bins at $z>4.1$.
    Black curves in each panel show the best-fit double power law,
    with the corresponding one-sigma (68.26\%) uncertainty shown by
    the grey shaded area.  There are 451, 270, and 69 qsos in each
    redshift bin from left to right, respectively.  These numbers are
    higher than those in Figure~\ref{fig:mosaic} because they include,
    respectively, 9, 7, and 3 qsos from \citet{2015AA...578A..83G}.
    The magnitude bins containing these qsos are shown in purple.  The
    red dashed curves show the double power law fits reported by
    \citet{2015AA...578A..83G} at $z=4.25, 4.75,$ and
    $5.75$.}
\end{figure*}

>>>>

\author[Kulkarni et al.]{{Girish Kulkarni$^{1}$\thanks{Email:
      kulkarni@ast.cam.ac.uk}, G\'abor Worseck$^{2}$ and Joseph
    F.~Hennawi$^{3}$} \\ $^1$Institute of Astronomy and Kavli
  Institute of Cosmology, University of Cambridge, Madingley Road,
  Cambridge CB3 0HA, UK \\ $^2$Institute f\"ur Physik und Astronomie,
  Universit\"at Potsdam, Karl-Liebknecht-Str.\ 24/25, 14476 Potsdam,
  Germany\\ $^3$Department of Physics, Broida Hall, UC Santa Barbara,
  Santa Barbara, CA 93106-9530 USA}
% Update Gabor's address to Potsdam
\date{Accepted ---. Received ---; in original form ---}

@@

%% Q notation is different on figure and in text.
%% MH15 curve seems wrong; it should reionize at z ~ 4 or so.
\begin{figure}
  \begin{center}
    \includegraphics[width=\columnwidth,keepaspectratio]{q.pdf}
  \end{center}
  \caption{Evolution of the \HeIII\ ionization fraction in our model,
    when the AGN luminosity function is integrated down to
    $M_{1450}=-21$ (blue curve) and $M_{1450}=-18$ (red curve).  The
    shaded regions show the one-sigma (68.26\%) uncertainty.  Other
    curves show models of \citet[solid grey]{2012ApJ...746..125H},
    \citet[dashed grey]{2015ApJ...813L...8M}, \citet[light blue and
      shaded region]{2016ApJ...828...90L},
    \citet[blue]{2018arXiv180104931P},
    \citet[brown]{2009ApJ...694..842M}, and
    \citet[green]{2014MNRAS.445.4186C}.  Note that we derive
    $Q_\nHeIII$ using Equation~(\ref{eqn:qHedot}), which ignores the
    presence of \HeII\ Lyman-limit systems and becomes inaccurate
    towards the end of \HeII\ reionization. \label{fig:qhe}}
\end{figure}

We now consider the implications of our AGN luminosity function models
for \HeII\ reionization.  We use the emissivity model developed in
Section~\ref{sec:e912} for this purpose.  The evolution of the volume-
averaged fraction $Q_\nHeIII$ of \HeIII\ is given by
\citep{2012ApJ...746..125H}
\begin{equation}
{dQ_\nHeIII\over dt}={\dot n_{\rm ion,4}\over \langle n_\nHe \rangle} -{Q_\nHeIII\over
\langle t_{\rm rec,He}\rangle},
\label{eqn:qHedot}
\end{equation}
where $\dot n_{\rm ion,4}$ is the number density of photons with
energy 4~Ry and above produced per unit time, $n_\nHe$ is the Helium
number density, and $\langle t_{\rm rec,He}\rangle$ is an average
recombination time scale for \HeIII.  We obtain the evolution of $\dot
n_{\rm ion,4}$ for AGN with $M_\mathrm{1450}<-21$ and
$M_\mathrm{1450}<-18$ by integrating the contribution of double power
law luminosity function obtained in distinct redshifts bins and
fitting \ref{eqn:e912fit} to the resultant \HeII-ionizing
emissivities.  We extrapolate the resultant fit to $z>7$, although the
contribution from this to \HeII-reionization is small.  The SEDs from
Equation~(\ref{eqn:sed}) and (\ref{eqn:sed_faint}) are assumed for
bright and faint AGN, respectively.  The recombination time scale in
Equation~(\ref{eqn:qHedot}) is given by
\begin{equation}
\langle t_{\rm rec}\rangle=[(1+2\chi) \langle n_\nH\rangle \alpha_B\,C]^{-1},
\label{eq:trec}
\end{equation}
where $n_\nH$ is the volume-averaged density of hydrogen, $\alpha_B$
is the case-B \HeIII\ recombination rate, and $\chi=0.083$ is the
cosmic number fraction of Helium.  We assume that the helium clumping
factor $C$ is given by \citep{2015ApJ...813L...8M}
\begin{equation}
  C = 2.8\left(\frac{1+z}{6}\right)^{-1.1}.
\end{equation}
We use the fitting function provided by \citet{1997MNRAS.292...27H}
for the recombination coefficient.  Note that
Equation~\ref{eqn:qHedot} neglects the \HeII\ Lyman-limit systems that
play an increasingly important role towards the end of Helium
reionization \citep{2009MNRAS.395..736B, 2017ApJ...851...50M}.  In the
absence of these self-shielded systems of high-density \HeII,
$Q_\nHeIII$ can continue to increase beyond unity.  We set it to one
when this happens.  This can be corrected by accounting for the
\HeII\ column density distribution and filtering the \HeII-ionizing
radiation background through it.  Unfortunately, the \HeII\ column
density distribution is itself uncertain.  We therefore consider our
treatment of $Q_\nHeIII$ accurate enough for the purpose of this
work.

The resultant reionization histories of \HeII\ are shown in
Figure~\ref{fig:qhe} for $M_\mathrm{1450}<-21$ and
$M_\mathrm{1450}<-18$ in red and blue, respectively.  We find that
Helium reionization is complete at $z\sim 3.2$ and $3.1$ in these
models, respectively, with a one-sigma uncertainty of about $\Delta
z\sim 0.2$.  This is likely consistent with the \HeII\ \lya effective
optical depth measurements of \citet{2016ApJ...825..144W}, who find
that the effective optical depth rises at $z>2.7$, but the incidence
of low optical depth sightlines beyond this redshift implies that the
substantial volume of \HeII\ was ionized already at $z=3.4$.
\HeII\ reionization begins at $z\gtrsim 5$ in our models.  This could
explain the transmissive sightlines reported by
\citet{2016ApJ...825..144W}, although a comparison with their data
requires a more detailed modelling of the IGM.

In Figure~\ref{fig:qhe}, we also show the reionization histories from
other models of \HeII\ reionization in the literature.  Helium
reionization is delayed in the model of \citet{2012ApJ...746..125H}
due to their reduced emissivity at $z\sim 3$.  In the
\citet{2015ApJ...813L...8M} model, a significant \HeII\ fraction is
ionized already at $z=6$ due to the very high emissivity assumed in
this model.  In the radiation-hydrodynamic simulations by
\citet{2016ApJ...828...90L}, the AGN population is calibrated to the
SDSS and COSMOS data at $z>3.5$ \citep{2012ApJ...755..169M,
  2013ApJ...768..105M}, combined with a model for quasar light curves.
The fiducial \HeII\ reionization history obtained by these authors is
shown in Figure~\ref{fig:qhe} along with an estimate of the
uncertainty.  Reionization ends considerably later in this model
($z\sim 2.6$).  Our model curves are closest in agreement to the
simulations of \citet{2009ApJ...694..842M} and the analytical model of
\citet{2018arXiv180104931P}.  Still, reionization completes earlier in
our model, partly due to the absence of \HeII\ Lyman-limit systems.
Our redshift of reionization agrees more closely with that in the
simulations presented by \citet{2014MNRAS.445.4186C}.

Helium reionization has also been studied by
\citet{2018MNRAS.473.1416M}, who constrained the evolution of the
emissivity from AGN using the effect optical depth measurement of
\citet{2016ApJ...825..144W}, combined with an analytical model of the
IGM.  While a detailed comparison with their work is difficult because
of reasons, \citet{2018MNRAS.473.1416M} find that the AGN emissivity
has to drop considerably rapidly at $z>4$. \citet{2018arXiv180109693K}
have recently presented a model of Helium reionization.  Reionization
ends at redshift 2.8 in their model.  The AGN luminosity function used
is by a combination of deteminations from various data sets
\citep{2015MNRAS.451L..30K}.  The integration limits are inconsistent
however as they are $0.01L_*$.  Resulting emissivity is comparable to
our model at $z\sim 2$ but drops more rapidly towards higher
redshifts.  For a steep EUV spectral index of $\alpha=-1.8$ their
helium reionization redshift is $z=2.8$.  \citet{2016arXiv160706467D}
considered the effect of AGN-dominated reionization on the Ly$\alpha$
opacity at $z>5$, He~\textsc{ii} Ly$\alpha$ opacity at $z\sim
3.1$--$3.3$, and the thermal history of the IGM.  In agreement with
\citet{2015MNRAS.453.2943C}, these authors found that AGN did provide
a plausible explanation for the large fluctuations in the Ly$\alpha$
opacity at $z>5$.  However, they found that reionization of
He~\textsc{ii} occurs much earlier in these AGN-dominated models (see
also \citealt{2018MNRAS.473.1416M}).

@@

We thank Eilat Glikman, Linhua Jiang, Nobunari Kashi\-kawa, Ian
McGreer, Nick Ross, Chris Willott, and Jinyi Yang for sharing data and
especially for sharing their quasar selection functions.  It is a
pleasure to acknowledge useful discussions with James Aird, Eduardo
Ba\~nados, Manda Banerji, Tirthankar Roy Choudhury, George Efstathiou,
Xiaohui Fan, Andrea Ferrara, Prakash Gaikwad, Martin Haehnelt, Paul
Hewett, David Hogg, Vikram Khaire, Sergey Koposov, Donald Lynden-Bell,
Roberto Maiolino, Richard McMahon, Daniel Mortlock, Ewald Puchwein,
Gordon Richards, Alberto Rorai, Bram Venemans and Stephen Warren.  GK
acknowledges support from ERC Advanced Grant 320596 `The Emergence of
Structure During the Epoch of Reionization'.
%% Joe, do you need to add anything to acknowledgements?

@@

\section{Comparison with G15}
\label{sec:conv}
%% Define 'G15' somewhere above.
In the double power law luminosity function models presented in
distinct redshift bins in Section~\ref{sec:bins}, we did not include
the 19 low-luminosity ($M_{1450}>-22.6$) AGN between redshifts $z=4.1$
and $6.3$ reported by G15.  While this was done in order to restrict
our sample to quasars with spectroscopic redshift determinations, it
is instructive to consider how our results are affected if the G15 AGN
are added to the analysis.  In their work, G15 found a shallower
faint-end slope ($\beta\sim -1.5$ to $-1.8$) for the luminosity
function at $4.1 < z < 6.3$, relative to our result from other AGN
samples at these redshifts ($\beta\sim -2.0$ to $-2.5$).  Still, G15
derived a higher 912\,\AA\ emissivity than our estimates
(cf.\ Figure~\ref{fig:e912}), so that in their analysis AGN can
produce all ionizing photons necessary to keep hydrogen ionized and
explain the \lya data.  The black points in
Figure~\ref{fig:params_giallongo} show the parameters of the double
power law luminosity function from our analysis of
Section~\ref{sec:bins}.  The red open circles show the parameter
values obtained when the G15 sample is added to the analysis.  We find
that the two results are highly consistent, showing that the G15
sample is consistent with our double power law fit obtained from other
AGN samples are comparable redshifts.  This is surprising as the
integrated 912\,\AA\ emissivities in our model are smaller than those
derived by G15.  Figure~\ref{fig:lf_giallongo} provides an
explanation.  As seen in this figure, the double power law fits
favoured by G15 (red dashed curves) are quite different from our fits
(black curves).  The characteristic luminosity $M_*$ obtained by G15
is much fainter ($\sim -23$ at $z = 5$) than that resulting out of our
analysis ($\sim -29$ at $z = 5$).  Thus the 912\,\AA\ emissivities are
enhanced in G15 because of the increase contribution from
intermediate-luminosity AGN in their model.
Figure~\ref{fig:lf_giallongo} suggests that this is possibly because
of the inclusion of SDSS Stripe 82 data from
\citet{2013ApJ...768..105M} and the AGN samples of
\citet{2010AJ....139..906W} and \citet{2015ApJ...798...28K} in our
analysis.  Additionally, the homogenisation of data in our work may
also cause part of the difference.

\begin{figure*}
  \begin{center}
    \includegraphics[width=0.7\textwidth]{evolution_g.pdf}
  \end{center}
  \caption{Effect of the 19 AGN reported by \citet{2015AA...578A..83G}
    on the best-fit double power law luminosity function parameters in
    redshift bins from $z=0$ to $7$.  Black points show parameter
    values from Figure~\ref{fig:evoln}.  Red open circles show the
    parameter values obtained when the sample of G15 is added to the
    analysis.  In both cases, vertical error bars show one-sigma
    (68.26\%) uncertainties, and horizontal error bars show widths of
    the redshift bins. \label{fig:params_giallongo}}
\end{figure*}

\begin{figure*}
  \begin{center}
    \includegraphics[width=\textwidth]{giallongo_compare.pdf}
  \end{center}
  \caption{Luminosity functions in three redshift bins at $z>4.1$.
    Black curves in each panel show the best-fit double power law,
    with the corresponding one-sigma (68.26\%) uncertainty shown by
    the grey shaded area.  There are 451, 270, and 69 AGN in each
    redshift bin from left to right, respectively.  These numbers are
    higher than those in Figure~\ref{fig:mosaic} because they include,
    respectively, 9, 7, and 3 AGN from G15.  The magnitude bins
    containing these AGN are shown in purple.  The red dashed curves
    show the double power law fits reported by G15 at $z=4.25, 4.75,$
    and $5.75$. \label{fig:lf_giallongo}}
\end{figure*}
